Conditional selectivity for triplestores

### 2012-09-17 conditional/trunk/selectivity-basic.tex
write down the equation parts that can be computed
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@370 43ef1190-90fd-4dd7-a769-d0121f02b80e

Finally, note that for variables, the selectivity will always be
unity, whether conditional or independent, since it will match all
possible RDF terms.

@@

In the follwing, we denote a bounded term with a lowercase letter, $s$
for a subject, $p$ for a predicate and $o$ for object, and variables
are denoted by a question mark, $?$. For example, $p =
\rdfterm{rdf:type}$. We also adopt SPLENDIDs convention of denoting
the number of all triples in certain data source $|d|$.

\subsection{Terms that can be computed}

\begin{equation}\label{eq:selpp}
\sel{P = p} = 
\end{equation}

@@

To relate this to VoID, we note that VoID includes the concept of
\emph{property partitions} (see section~4.5 in \cite{voidnote}): ``A
property-based partition contains only those triples of a dataset that
use a particular predicate.'', and since a partition is itself a
dataset, the same statistical properties can be applied, such as
\rdfterm{void:triples}, \rdfterm{void:distinctSubjects} and
\rdfterm{void:distinctObjects}.

>>>>

Finally, note that for variables, the selectivity will always be
unity, since it will match all possible RDF terms.

@@

In the following, we shall have to introduce to more elaborate
notation to make sure we maintain control of bounded and unbounded
terms. We denote a bounded term with a lowercase letter, $s$
for a subject, $p$ for a predicate and $o$ for object, and variables
are denoted by a question mark, $?$. For example, $p =
\rdfterm{rdf:type}$. We also adopt SPLENDIDs convention of denoting
the number of all triples in certain data source $|d|$.

\subsection{Parts that can be computed}

There will be a number of different equations that arises from using
the commutative property of intersections, and some parts of these
equations is possible to compute from VoID descriptions. We have
already noted the most trivial case, for variables we have:

\begin{equation}\label{eq:selvar}
\sel{S = ?} = \sel{P = ?} = \sel{O = ?} = 1 .
\end{equation}

We note that VoID includes the concept of \emph{property partitions}
(see section~4.5 in \cite{voidnote}): ``A property-based partition
contains only those triples of a dataset that use a particular
predicate.'', and since a partition is itself a dataset, the same
statistical properties can be applied, such as \rdfterm{void:triples},
\rdfterm{void:distinctSubjects} and \rdfterm{void:distinctObjects}.

In the case where the predicate is known, we have
\begin{equation}\label{eq:selPp}
 \sel{P = p} =
   \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|} ,
\end{equation}
where the notation is inspired by SPARQL 1.1 Property Paths, for
example the relevant part of a VoID description for the second example
in section~\ref{sec:example} would be:
\begin{verbatim}
 </dataset> a void:Dataset ;
            void:propertyPartition [
                void:property foaf:firstName;
                void:triples 2 .
            ]
\end{verbatim}
where $p = \rdfterm{foaf:firstName}$.

In the case where there is a given predicate but the subject is a
variable, we can use the \rdfterm{void:distinctSubjects} property:
\begin{equation}\label{eq:selSvPp}
 \sel{S = ? | P = p} =
   \frac{\rdfterm{void:propertyPartition/void:distinctSubjects}}{\rdfterm{void:propertyPartition/void:triples}} ,
\end{equation}
and similarly where the object is a variable:
\begin{equation}\label{eq:selOvPp}
 \sel{O = ? | P = p} =
   \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
\end{equation}

We also have in the special case of a given class, i.e. predicate
\rdfterm{rdf:type} or $P = a$ and $O = c$, we can use
\rdfterm{void:classPartition} (also defined in section~4.5 in
\cite{voidnote}):
\begin{equation}\label{eq:selSvPaOc}
 \sel{S = ? | P = a \cap O = c } =
    \frac{\rdfterm{void:classPartition/void:triples}}{|d|} .
\end{equation}

The above are the parts of the equations that can be computed directly
from the VoID description, and in the next section we shall see how
they can be exploited most efficiently.

@@
